### 2019-02-25 paper/intro-notes.tex
Add paragraphs on spacing and health outcomes

[emphasize that birth spacing is a superset of DSB?]
Of more direct interest here is, however, the potential for differential \emph{spacing} 
behavior, 
both because it has been used as a measure for son preference and because birth spacing 
may directly impact health and other outcomes for both children and mother.
In India the sex of the last-born child significantly impact the duration to next birth,
with the expected birth interval about three percent longer if the last-born was son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are very small \citep{Pong1994}.
}
India is not alone in showing an effect of sex composition on subsequent birth spacing.
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia the evidence is more mixed.
North Africa shows shorter spacing in the absense of sons, while a similar pattern does
not exists in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

@@

>>>>

[emphasize that birth spacing is a superset of DSB?]
Of particular interest here is, however, the potential for differential \emph{spacing} 
behavior, 
both because it has been used as a measure for son preference and because birth spacing 
may directly impact health and other outcomes for both children and mother.
In India the sex of the last-born child significantly impact the duration to next birth,
with the expected birth interval about three percent longer if the last-born was son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are very small \citep{Pong1994}.
}
India is not alone in sex composition affecting subsequent birth spacing.
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia the evidence is more mixed.
North Africa shows shorter spacing in the absense of sons, while a similar pattern does
not exists in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

There is an even more substantial literature addressing the potential effects of birth 
spacing on health outcomes, but, although most of the research finds a negative impact of 
short spacing on health, identifying the causal mechanisms have proven more difficult
\citep{Conde-Agudelo2006,Conde-Agudelo2012}.
Broadly speaking, there are three pathways that birth intervals can affect health.
First, maternal depletion, where the mother does not have sufficient
time to recover after a pregnancy.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiency.
}
Second, disease transmission both from mother to child and between children close in age.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is especially of interest when spacing decisions are driven by 
son preference.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to negatively affect the older child.

In India, there is evidence that both the prior and the subsequent child have higher
likelihood of dying the closer they are spaced, although the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Maitra2008,Makepeace2008}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explain a limited amount
of neonatal mortality.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, and El Salvadore that
short intervals between birth has negative health effects
\citep{Curtis1993,Davanzo2008,Gribble2009,Saha2013}.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again quicker, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.

@@

What is common in the literature reviewed above it that the potential effects of 
sex selection on spacing has not been incorporated.
